The language of generalization: Probability, vagueness, and interaction

### 2018-02-18 paper/coverletter.tex
streamline presentation in first few pagraphs, lightly edit throughout

We enclose the manuscript \emph{The Language of Generalization} to be considered for publication in \emph{Psychological Review}. 
This paper explores the linguistic means by which generalizations are communicated.  
We present a Bayesian model of endorsing generalizations and show that it predicts context-sensitive human judgments with high quantitative accuracy across several experiments in diverse domains.
The theory suggests that the language of generalizations can be understood in terms of a simple semantic theory that operates over beliefs, represented as probability.

Understanding how the world works requires learning generalizations. 
The kinds of observations that would facilitate such learning, however, may be rare (e.g., figuring out that lightning strikes tall objects) or costly (e.g., learning that certain plants are poisonous).
We do not need to need taste every plant to figure out which ones are poisonous, however.
Language lets us learn generalizations from each other.

The language of generalization is most commonly studied in predications of categories (e.g., ~\emph{Birds fly}), otherwise known as \emph{generic language}.
But different semantic types can be described in generalization, including events (\emph{habitual language}; e.g., ~\emph{Mary swims after work}) and causal relationships (\emph{causal language}; e.g., ~\emph{Staring at the sun makes you go blind}).
Generalizations are ubiquitous in everyday conversation (\emph{Movie theaters are cold.}), the speech of young children (\emph{Slides aren't for grown ups!}), child-directed speech (\emph{Dogs are friendly.}), political and scientific discourse (\emph{Politicians make promises.}, \emph{Psychology experiments don't replicate.}), stereotypes (\emph{Boys are good at math.}), motivation (\emph{Big girls eat broccoli!}), and many other facets of experience.

%Given that inductive generalization can be described by probability (e.g., the probability that the next instance of the category will have a feature), it seems natural that the core meaning generalizations in language would relate to this predictive probability.
Despite their ubiquity and morphosyntactic simplicity, the language of generalization has a meaning, which is philosophically puzzling and has resisted precise formalization. 
%However, the language of generalizations displays extreme sensitivities to context, which has heretofore crippled previous attempts to formalize this major function of language. 
%Though this type of language is found everywhere, the meaning of generalization is philosophically puzzling and has resisted precise formalization. 
%The major issue in formalizing the language of generalizations is determining what makes such statements true or false.
For example, generics can describe properties contained by a majority (e.g., \emph{Dogs bark}) but still can be felicitous based on only a minority with the property (e.g., \emph{Robins lay eggs}, even though only adult, female, fertile robins do). 
\emph{Mosquitos carry malaria} sounds true, even though only 1\% of them do, highlighting that the acceptability of generics can be so flexible as to include cases where only the weakest quantifier (i.e., \emph{some}) would suffice. 
Habitual and causal language display similar context-sensitivities:
If John smokes once a year, one might hesitate to endorse that he \emph{smokes}.
Yet, climbing a mountain once a year makes John really seem like he \emph{climbs mountains}.
\emph{Drinking moonshine makes you go blind} and so does \emph{staring at the sun}, but neither statement conveys precise information about the relative risk.

@@

%The dominant view in psychology has been to focus on special conceptual relations that supposedly determine what makes the language the generalization true or false.
In this paper, we take the context-sensitivity of generalizations seriously and formalize the hypothesis that the core meaning of generalizations is \emph{simple but underspecified}. 
We embed this idea in a simple, rational model of communication and use it to address the fundamental question: What makes generalizations in language true or false?
Our answer is that endorsement (a true or false judgment) is a speaker's decision about whether or not to say the generalization to a naive listener.
Our model is the first formal model to make precise quantitative predictions about the language of generalization.

We have previously received peer-review of a related manuscript submitted to top-ranking journals including \emph{Psychological Review}.
Those reviews raised a umber of points regarding the presentation and generality of our work at the time, and highlighted the need for comparison to other models. 
The current manuscript differs from previous submissions a number of important ways.
First, we have completely re-written the paper for clarity and focus, especially simplifying and streamlining the modeling sections.
Second, we have extended from generalizations about categories, or generic language, to investigate three case studies: generalizations about categories, events, and causes.
We are able to explain endorsements across these three domains without adopting domain-specific assumptions. 
Furthermore, each case study naturally lends methodological advantages to provide further tests of our model. (E.g. with causal generalizations, we are able to \emph{manipulate} (rather than merely measure) prior knowledge.)
%The empirical and much of the philosophical literatures have focused on generalizations about categories, or generic language.
%Our previous manuscript also focused on categories; in this paper, however, we investigate three case studies: generalizations about categories, events, and causes. 
%Immediately we see the generality of our formal model: We are able to explain context-sensitive endorsements across these three domains without adopting domain-specific assumptions.
%Furthermore, each case study naturally lends methodological advantages to provide further tests of our model.
%With event generalizations, we are able to interrogate the underlying semantic scale, and find that speakers' subjective beliefs about what is likely to be the case in the future, and not the objective truth about what has been the case in the past, is what drives endorsements. 
%With causal generalizations, we are able to \emph{manipulate} (rather than merely measure) prior knowledge, and show it is in fact causally related to endorsement. 
Finally, we develop three alternative models: two linear models based on empirically-elicited statistical knowledge and one rich, cognitive model which lacks the \emph{underspecification of meaning} at the core of our hypothesis. 
These alternative models fail to account for key aspects of the context-sensitivity that is observed in the data and that our full model predicts.
%These alternative models and the empirical paradigms used provide strong tests of the theoretical contribution of our paper. 
%This novel approach for generalizations in language opens up a number of interesting questions about the nature of conceptual structure and its linguistic traces, which we discuss at the end of the paper.
Overall we feel that this paper represents a much improved, generalized and streamlined, study of generalization in language.
(We would be happy to provide detailed responses to previous reviewers, however we feel this would be of limited use given the major changes in content and presentation.)

@@

This work will be of interest to psychologists, as well as linguists, philosophers, and computer scientists interested in language and interaction.
The puzzles of generalizations in language are immediately compelling to a general audience, the more so because of their connections to cognitive development, cultural transmission of knowledge, and propagation of stereotypes.
Our solution represents the first formal model of generalization in language to be validated by quantitative agreement with human behavioral data.
It is not, however, uncontroversial: we champion a statistical view that has been often dismissed in the literature.
Indeed, in presentations around the world we have found this work to excite broad interest and engagement.

>>>>

We enclose the manuscript \emph{The Language of Generalization} to be considered for publication in \emph{Psychological Review}. 
This paper explores the linguistic means by which generalizations are communicated.  
We present a Bayesian model of endorsing generalizations and show that it predicts context-sensitive human judgments with high quantitative accuracy across several experiments in diverse domains.
The theory suggests that the language of generalization can be understood in terms of a simple semantic theory that operates over beliefs, represented as probability.

Understanding how the world works requires learning generalizations. 
The kinds of observations that would facilitate such learning may be rare, however (e.g., figuring out that lightning strikes tall objects) or costly (e.g., learning that certain plants are poisonous).
We do not need to need taste every plant to figure out which ones are poisonous, however.
Language lets us learn generalizations from each other.

The language of generalization is most commonly studied in predications of categories (e.g.,~``Birds fly.''), otherwise known as \emph{generic language}.
But generalization exists beyond classically defined categories (e.g., natural and artifact kinds): Events are described in generalization (e.g.,~``Mary swims after work.''; \emph{habitual language}) as well as causal relationships (e.g.,~``Staring at the sun makes you go blind''; \emph{causal language}).
Generalization are ubiquitous in everyday conversation (``Movie theaters are cold.''), the speech of young children (``Slides aren't for grown ups!''), child-directed speech (``Dogs are friendly.''), political and scientific discourse (``Politicians make promises.'', ``Psychology experiments don't replicate.''), stereotypes (``Boys are good at math.''), motivation (``Big girls eat broccoli!''), and many other facets of experience.

%Given that inductive generalization can be described by probability (e.g., the probability that the next instance of the category will have a feature), it seems natural that the core meaning generalizations in language would relate to this predictive probability.
Despite their ubiquity and morphosyntactic simplicity, the language of generalization has a philosophically puzzling meaning, which has resisted precise formalization. 
%However, the language of generalizations displays extreme sensitivities to context, which has heretofore crippled previous attempts to formalize this major function of language. 
%Though this type of language is found everywhere, the meaning of generalization is philosophically puzzling and has resisted precise formalization. 
%The major issue in formalizing the language of generalizations is determining what makes such statements true or false.
For example, true generic statements can describe properties contained by a majority of the category (e.g., ``Dogs bark.''), roughly half of the category (e.g., ``Robins lay eggs.'' is only true of adult, female, fertile robins), and even an outstanding minority (e.g., ``Mosquitos carry malaria.'' is true of less than 1\% of mosquitoes). 
Habitual and causal language display similar context-sensitivities:
``John smokes'' may be infelicitous if he does so only once a year, but ``Mary climbs mountains'' seemingly is fine with the same frequency.
Both \emph{drinking moonshine}  and \emph{staring at the sun} will ``...make you go blind'', though the latter seems more dangerous than the former despite neither statement conveying precise risk information.

@@

%The dominant view in psychology has been to focus on special conceptual relations that supposedly determine what makes the language the generalization true or false.
In this paper, we take seriously the context-sensitivity of such language and investigate the hypothesis that the meaning at the core of the language of generalization is \emph{simple but underspecified}. 
We formalize and embed this idea in a simple, rational model of communication and use it to address the fundamental outstanding question: What makes generalizations true or false?
Our answer is that endorsement (a true or false judgment) is a speaker's decision about whether or not to say the \emph{simple but underspecified} generalization to a naive listener.
Our model is the first formal model to make precise quantitative predictions about human understanding of generalizations in language.

We have previously received peer-review of a related manuscript submitted to top-ranking journals including \emph{Psychological Review}.
Those reviews raised a umber of points regarding the presentation and generality of our work at the time, and highlighted the need for comparison to other models. 
The current manuscript differs from previous submissions in a number of important ways.
First, we have completely re-written the paper for clarity and focus, especially simplifying and streamlining the modeling sections.
Second, we have extended from generalizations about categories (or \emph{generic language}) to investigate three case studies: generalizations about categories, events, and causes.
We are able to explain endorsements across these three domains with minimal domain-specific assumptions. 
Furthermore, each case study naturally lends methodological advantages to provide further tests of our hypothesis/model (e.g., with causal generalizations, we are able to manipulate---rather than merely measure---prior knowledge).
%The empirical and much of the philosophical literatures have focused on generalizations about categories, or generic language.
%Our previous manuscript also focused on categories; in this paper, however, we investigate three case studies: generalizations about categories, events, and causes. 
%Immediately we see the generality of our formal model: We are able to explain context-sensitive endorsements across these three domains without adopting domain-specific assumptions.
%Furthermore, each case study naturally lends methodological advantages to provide further tests of our model.
%With event generalizations, we are able to interrogate the underlying semantic scale, and find that speakers' subjective beliefs about what is likely to be the case in the future, and not the objective truth about what has been the case in the past, is what drives endorsements. 
%With causal generalizations, we are able to \emph{manipulate} (rather than merely measure) prior knowledge, and show it is in fact causally related to endorsement. 
Finally, we develop three alternative models: two linear models based on empirically-elicited statistical knowledge and one rich, cognitive model which lacks the \emph{underspecification of meaning} at the core of our hypothesis. 
These alternative models fail to account for key aspects of the context-sensitivity that is observed in the data and that our full model predicts.
%These alternative models and the empirical paradigms used provide strong tests of the theoretical contribution of our paper. 
%This novel approach for generalizations in language opens up a number of interesting questions about the nature of conceptual structure and its linguistic traces, which we discuss at the end of the paper.
Overall we feel that this paper represents a much improved, generalized and streamlined investigation of generalization in language.
(We would be happy to provide detailed responses to previous reviewers; presently, we feel this would be of limited use given the major changes in content and presentation.)

@@

This work will be of interest to psychologists, as well as linguists, philosophers, and computer scientists interested in language and interaction.
The puzzles of generalizations in language are immediately compelling to a general audience, the more so because of their connections to cognitive development, propagation of stereotypes, and cultural transmission of knowledge.
Our solution represents the first formal model of generalization in language to be validated by quantitative agreement with human behavioral data.
It is not, however, uncontroversial: we champion a statistical view that has been often dismissed in the literature.
Indeed, in presentations around the world we have found this work to excite broad interest and engagement.
